Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-17 paper/main.tex
format toc
add todo
add comments from christian
Co-authored-by: Christian Fleck <christian.fleck@fdm.uni-freiburg.de>

%###################################################################################################
\begin{abstract}
    \aclp{abm} have become popular tools in the study of complex cellular systems.
    However, despite their popularity, it is often difficult to systematically estimate their
    parameters and perform advanced analyses e.g. using the profile-likelihood.
    In this work, we study the mechanical properties of rod-shaped bacteria and provide techniques
    for estimating their parameters and performing identifiability analysis on them.
    Our approach follows a blueprint of a modular framework which can be adapted to other research
    questions.
    We use time series of microscopic images as input data and incorporate advanced effects as the
    elasticity of the rod.
    The developed model serves as a generalized foundation to describe multiple species of
    rod-shaped bacteria such as \textit{E.Coli} or \textit{B.Subtilis}.
\end{abstract}

@@

\pagebreak
\renewcommand{\contentsname}{Table of Contents (remove before submission)}
\tableofcontents
\pagebreak

%###################################################################################################
\section{Introduction}

\acp{abm} have proven to be powerfull tools in the study of complex cellular systems.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

Despite their popularity, \acp{abm} face multiple challenges in practise.
Constructing a fully working \ac{abm} simulation from scratch is an intricate process that requires
a large set of skills and time.
It is therefore common practise to reuse existing simulation tools, which brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability \textbf{TODO CITATION}.
Typically researchers aim to estimate every individual parameter of their model.
This is achieved using methods such as the profile-likelihood and model reduction
techniques \textbf{TODO CITATION}.
When it comes to \acp{abm} however, it is often only shown that the results which have been
generated by an \ac{abm} lead to the desired phenomenon which aligns with the observed biological
reality.
But this process does not actually estimate the necessary parameters directly \textbf{TODO CITATION}
and reduces the confidence in the constructed models and raises concerns with respect to the
validity of the obtained results.

In this work, we address these challenges using a general-purpose mechanical model for rod-shaped
bacteria.
Our approach provides various methods which can be combined to systematically estimate the
parameters of our model.
We deliberately distinguish between conceptually separate components within this framework.
This modularity highlights a "blueprint" which can be adapted and applied to other \acp{abm}.

@@

This work adresses some of these limitations.
We use microscopic image time series as inputs to obtain information about the spatial
configuration of the cells.
This information can be used as initial values for a numerical simulation.
By varying the parameters of the model and comparing the results to the microscopic data, we can
estimate and quantify the parameters of our model.
Our approach takes a general stance which incorporates effects that are observed within
\textit{B.Subtilis} and \textit{E.Coli}.

@@

@@

The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
In this work, we aim to establish a model which is abstract enough such that it can be reused for
many types of rod-shaped bacteria but also explicit enough such that we can quantify its parameters.
We are thus aiming to work with a set of assumptions that is as general as possible but still
specific to rod-shaped bacteria.
With this delicate balance in mind, it is clear that many cellular processes which might be specific
to one particular celltype or simply not necessary to assume in order to construct our model, will
be omitted alltogether.
This concerns effects such as polar interactions~\cite{Duvernoy2018}, extracellular
reactions~\cite{Wang2010} or differentiation processes~\cite{vanGestel2015,Lpez2010} and many other.
When describing our system, we distinguish between cellular aspects which ocurr within a single
cell and cell-cell interactions between numerous bacteria.
We restrict ourselves to the aspects presented in table~\ref{table:simulation-aspects}.
For a collection of other known effects which could be considered, we refer to the supplementary
section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}.

\paragraph{(C) Cellular}
It was long believed that bacteria lack cytoskeletal filaments but with the end of the last century,
it was discovered that the MreB protein fills this role since it “[..] forms an actin-like
cytoskeleton in bacteria [..]”~\cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments~\cite{Dersch2020}.
The effect of MreB on the shape of cells during growth was shown in more detail
by~\cite{Ursell2014}.
They found that MreB localizes specifically at regions of negative curvature and thus introduces
heterogeneous cell growth which results in a bending force during the growth of cells.
It also means that the cells are able to sense the curvature of the cell wall through cytoskeleteal
MreB filaments during growth.
\cite{Bratton2018} showed that the trans-membrane protein RodZ can modulate MreB density and
facilitates local MreB assembly.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment.
These results were also confirmed by the studies of~\cite{Wang2010}.
In their experiments, a single bacterium was fixed at the bottom pole while an optically trapped
bead was placed on the upper pole.
They could now exert a force on the free part and thus study the response of the cell under
different conditions.
When treating cells with A22, an antibiotic which disassembles
MreB~\cite{IWAI2002,Gitai2005,Karczmarek2007,Bean2009}, a higher displacement with identical force
could be measured, thus indicating that MreB increases mechanical rigidity of the cell.

While some rod-shaped bacteria grow by inserting new material at the ends, \textit{E.Coli} and
\textit{B.Subtilis} utilize a method where new material is inserted along the cylindrical part of
the rod.
The details of the division event (i.e. Z-Ring formation) are very intricate and are heavily
studied as a subject on their own~\cite{Harry2001}.
On a higher level view, the rod is split apart at a point close to the center of the cell, thus
creating two new daughter cells.
We know that the best predictor for when the division event occurs, is the length of the rod as
opposed to the age of bacterium~\cite{Robert2014}.
Furthermore, only the growth of the whole colony but also the individual growth of the bacteria is
exponential~\cite{Amir2014,Takeuchi2005}.

@@

All these considerations become even more intricate when not considering interactions of bacteria
with surfaces but bacteria with each other.
Gram-negative bacteria (eg. E.Coli) express outer membrane proteins (adhesins) which allow them to
attach to various hosts (Beachey 1981, Vaca et al. 2019).
In addition, the flagella can play an important role in the direct adhesion between
cells~\cite{Haiko2013}.
Thus it is not correct to blindly apply the results for adhesion to surfaces to the interactions
between bacteria.

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod into two new agents.\\
            % Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (4) & Individual Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (5) & Adhesion &
            Bacteria attract at longer distances and adhere to each other at shorter
            distances~\cite{Verwey1947,Trejo2013}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{
        This table lists aspects of cellular behaviour which we consider in our computational model.
        It is split into cellular aspects (C) which concern only an individual bacterium and
        interactions between cells (CC).
        These aspects are generalized in order to fit to as many types of rod-shaped bacteria as
        possible.
        These are the assumptions for the basic model that we are presenting here.
        In order for our methods to work, we are really only required to satisfy the assumption of
        our spatial representation (see section~\ref{subsec:spatial-representation}) and thus, these
        assumptions can also be loosened if required.
    }
\end{table}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation}
\label{subsec:spatial-representation}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.

With these fundamental thoughts layed out, we retain the flexibility to represent many models such
as varying shapes since we do not have to make any assumptions on the type of interactions between
two bacteria.
Even concerning the shape of bacteria, this assumption still allows us to mdoel elliptical cells by
modifying the interaction type as well as other effects such as polarity of
interactions~\cite{Duvernoy2018}.
Furthermore, we also did not have to assume any particular mechanical dynamics of the agents, only
their spatial representation, thus leaving in particular room for plastical or elastic deformations
of the rod and other more intricate mechanisms.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ have been exaggerated such that they can be identified more easily.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with $0$ curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Mechanics}
\label{subsection:mechanical-model-mechanics}

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness and flexibility of the bacterial rod~\cite{Amir2014_2}.
As explained previously, MReB will counteract the bending of the rod, resulting in a stiffening
force.
In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988}
but in order to construct a model which can be parametrized and by method of Occam's Razor, we chose
to model the bacterium as a bending beam, thus utilizing the simplest solution which can still
produce the desired effects.
Such a beam can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
springs of lengths $l_i$.
We use the notation $\vec{c}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

@@

\begin{align}
    \partial_t^2 \vec{x}_i &= \partial_t\vec{x}_i + \sqrt{2D}\vec{\xi}_i\\
    \partial_t\vec{x}_i &= \vec{F}_{i,\text{total}} - \lambda \partial_t\vec{x}_i
\end{align}

where $D$ is the diffusion constant, $\vec{\xi}$ is the wiener process and $\lambda$ a damping
constant.
Together they determine the stochastic motion of the individual vertices which can arise from
recurring collisions with small particles of the surrounding medium or other minor fluctuations.

@@

When calculating forces which act between the cells, we can use a simplified model to circumvent the
numerically expensive integration over the complete length of the rod.
Given a vertex $\vec{x}_i$ on one cell, we calculate the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{w}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

@@

The Mie~\cite{Mie1903} potential is the generalization of the well-known
Lennard-Jones~\cite{Jones1924} potential.
It was also designed to describe the interactions of molecules but due to the flexible nature of its
parameters, can be easily adapted to other situations.
It is able to describe a wider range of shapes compared to the Morse potential but is also stiffer
and not bound from above.
In our simulation, we enforce an upper bound to guarantee numerical regularity.
It has 4 parameters and is given by
\begin{align}
    U(r) &= C\epsilon\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}\\
    V(r) &= \min(U(r), \beta)\theta(r-\zeta)
\end{align}
where $C$ is a constant that depends on the two exponents $n,m$ that describe the overall shape of
the potential.
The strength is given by $\epsilon$ and the radius is encoded in $\sigma =
(m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the radii of the two interacting particles.

@@

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$.
In most of our simulations, we choose to use identical lengths for all segments (i.e. $l_i=l$).
However, in order to generalize our approach we assume that individual lengths of segments could
vary.
Our implemented model allows us to reduce the growth rate with increasing number of neighbors
$n\leq N$.
This option provides a phenomenological approach for modeling slower growth in larger colonies but
is be toggled off by default thus removing the additional term in brackets of
equation~\ref{eq:growth-rod-lengths}.

@@

@@

@@

%###################################################################################################
\section{Application 2: \textit{E.Coli.} growing on Agar}
\label{section:application-2}

>>>>

%###################################################################################################
\begin{abstract}
    \aclp{abm} have become popular tools in the study of complex cellular systems.
    However, despite their popularity, it is often difficult to systematically estimate their
    parameters and perform advanced analyses, e.g., using the profile-likelihood.
    In this work, we study the mechanical properties of rod-shaped bacteria and provide techniques
    for estimating their parameters and performing identifiability analysis on them.
    Our approach may serve as a blueprint for a modular framework which can be adapted to other
    research questions.
    We use time series of microscopic images as input data and incorporate advanced effects as the
    elasticity of the rod.
    The developed model serves as a generalized foundation to describe multiple species of
    rod-shaped bacteria such as \textit{E.Coli} or \textit{B.Subtilis}.
\end{abstract}

@@

\pagebreak
\renewcommand{\contentsname}{Table of Contents (remove before submission)}
\tableofcontents
\vfill
\pagebreak

%###################################################################################################
\section{Introduction}
\textbf{
    Mein Vorgehen: ich passe Formulierungen an, wenn ich im Prinzip mit der Aussage einverstanden
    bin, aber den Wortlaut verbesserungswürdig finde.
    Ich schreibe Kommentare, dort wo ich eher grundsätzliche Schwierigkeiten sehe.
    Dort wo ich nichts geändert habe und keinen Kommentar gegeben habe, kann ich zu einem späteren
    Zeitpunkt doch Änderungen vornehmen wollen.
}

\acp{abm} have proven to be powerfull tools in the study of complex cellular
systems~\cite{Pleyer2023}.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells \textbf{TODO CITATION}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

Despite their popularity, \acp{abm} face multiple challenges in practise.
The construction of a fully functional \ac{abm} simulation from the ground up constitutes an
intricate and laborious process, necessitating a comprehensive array of skills. It is therefore
common practise to reuse existing simulation tools, which, in turn, brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability \textbf{TODO CITATION}. Typically researchers estimate the parameter of their model
using methods such as Maximum Likelihood or Maximum a Posteriori.

[Gabor, Attila, and Julio R. Banga. ‘Robust and Efficient Parameter Estimation in Dynamic Models of Biological Systems’. BMC Systems Biology 9, no. 1 (2015): 74. https://doi.org/10.1186/s12918-015-0219-2., 
Banga, Julio R, and Eva Balsa-Canto. ‘Parameter Estimation and Optimal Experimental Design’. Essays in Biochemistry 45 (2008): 195–209. https://doi.org/10.1042/BSE0450195., 
Ashyraliyev, Maksat, Yves Fomekong-Nanfack, Jaap A Kaandorp, and Joke G Blom. ‘Systems Biology: Parameter Estimation for Biochemical Models’. FEBS Journal 276, no. 4 (2009): 886–902. https://doi.org/10.1111/j.1742-4658.2008.06844.x.] \\

check for parameter identifiability using methods like profile-likelihood or apply model reduction techniques \\

[Kreutz, Clemens, Andreas Raue, Daniel Kaschek, and Jens Timmer. ‘Profile Likelihood in Systems Biology’. The FEBS Journal, May 2013, n/a-n/a. https://doi.org/10.1111/febs.12276.
Raue, Andreas, Johan Karlsson, Maria Pia Saccomani, Mats Jirstrand, and Jens Timmer. ‘Comparison of Approaches for Parameter Identifiability Analysis of Biological Systems.’ Bioinformatics (Oxford, England), ahead of print, January 2014. https://doi.org/10.1093/bioinformatics/btu006.
Simpson, Matthew J. ‘Data-Informed Model Reduction for Inference and Prediction from Non-Identifiable Models’. Preprint, Synthetic Biology, 21 March 2025. https://doi.org/10.1101/2025.03.20.644462.]\\

In the context of \acp{abm}, however, it is often only shown that the generated results lead to the
desired phenomenon aligning with the observed biological reality.
But this process does not actually estimate the necessary parameters directly \textbf{TODO CITATION}.
Consequently, it reduces the confidence in the constructed models and raises concerns with respect
to the validity of the obtained results.
In this work, we address these challenges using a general-purpose mechanical model for rod-shaped
bacteria as an example.
Our approach provides various methods which can be combined to systematically estimate the
parameters of our model. We deliberately distinguish between conceptually separate components within
this framework.
This modularity highlights a "blueprint" which can be adapted and applied to other \acp{abm}.

@@

\textbf{[Was mir hier nicht gefällt, ist, dass du von Parameter schätzen bei ABMs kommend ziemlich detailliert über Mechanik redest. Das passt nicht. Man erwartet mehr Infos über den Blueprint und ist verwirrt. Worum geht es hier?]}

This work addresses some of these limitations.
We use microscopic image time series as inputs to obtain information about the spatial configuration
of the cells.
This information can be used as initial values for a numerical simulation.
By varying the parameters of the model and comparing the results to the microscopic data, we can
estimate and quantify the parameters of our model.
Our approach takes a general stance which incorporates effects that are observed within
\textit{B.Subtilis} and \textit{E.Coli}.

@@

\textbf{[Ich denke nicht, dass $\vec{x}_i$ hier nötig ist. Es bietet keinerlei Information. Du kannst alles so schreiben ohne das mathematische Symbol. Du musst den Bogen hinbekommen von dem allgemeinen Problem des Parameterschätzen bei ABMs zu dem spezifischen Problem, welches du hier verhandelst.]}

@@

We will go into the details of the model for rod-shaped bacteria.
The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
In this work, we aim to establish a model which is abstract enough such that it can be reused for
many types of rod-shaped bacteria but also explicit enough such that we can quantify its parameters.
We are thus aiming to work with a minimal set of assumptions that is as general as possible but
still specific to rod-shaped bacteria.

Given this delicate balance, it is evident that numerous cellular processes will be excluded due to
their specificity to particular cell types or their irrelevance to the construction of our model.
This encompasses a range of effects, including polar interactions~\cite{Duvernoy2018}, extracellular
reactions~\cite{Wang2010} or differentiation processes~\cite{vanGestel2015,Lpez2010}, among numerous
others.
When describing the bacterial system, we distinguish between intra-cellular processes and cell-cell
interactions.
We restrict ourselves to the features discussed below and summarized in
Table~\ref{table:simulation-aspects}.
For a collection of other known aspects which could be considered, we refer to the supplementary
section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}.

\paragraph{(C) Cellular}
In bacteria the MreB protein forms an actin-like cytoskeleton~\cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments~\cite{Dersch2020}.
The effect of MreB on the shape of cells during growth was shown in more detail
by~\cite{Ursell2014}. They found that MreB localizes specifically at regions of negative curvature
and thus introduces
heterogeneous cell growth which results in a bending force during the growth of cells.
It also means that the cells are able to sense the curvature of the cell wall through cytoskeleteal
MreB filaments during growth.
\cite{Bratton2018} showed that the trans-membrane protein RodZ can modulate MreB density and
facilitates local MreB assembly.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment. These results were also confirmed by the studies of~\cite{Wang2010}.
In their experiments, a single bacterium was fixed at the bottom pole while an optically trapped
bead was placed on the upper pole. They could now exert a force on the free part and thus study the
response of the cell under different conditions. When treating cells with A22, an antibiotic which
disassembles MreB~\cite{IWAI2002,Gitai2005,Karczmarek2007,Bean2009}, a higher displacement with
identical force could be measured, thus indicating that MreB increases mechanical rigidity of the
cell.

While some rod-shaped bacteria grow by inserting new material at the ends, \textit{E.Coli} and
\textit{B.Subtilis} utilize a method where new material is inserted along the cylindrical part of
the rod. The details of the division event (i.e. Z-Ring formation) are very intricate and are heavily
studied as a subject on their own~\cite{Harry2001}.
On a higher level view, the rod is split apart at a point close to the center of the cell, thus
creating two new daughter cells.
We know that the best predictor for when the division event occurs, is the length of the rod as
opposed to the age of bacterium~\cite{Robert2014}.
Furthermore, not only the growth of the whole colony but also the growth of individual bacteria is
exponential~\cite{Amir2014,Takeuchi2005}.
\textbf{[Bitte etwas kürzen, da du hier keinen Review schreibst und dich deswegen um Prägnanz bemühen musst.]}

@@

All these considerations become even more intricate when interactions between bacteria are considered.
Gram-negative bacteria (eg. E.Coli) express outer membrane proteins (adhesins) which allow them to
attach to various hosts (Beachey 1981, Vaca et al. 2019).
In addition, the flagella can play an important role in the direct adhesion between
cells~\cite{Haiko2013}.
Thus it is not correct to blindly apply the results for adhesion to surfaces to the interactions
between bacteria.

\textbf{[Man versteht nicht so recht, warum du das alles hier erzählst und was daraus folgt. Die Verbindung zu deinem Modell fehlt.]}

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod into two new agents.\\
            % Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (4) & Process Parameters &
            We take a potential heterogeneity of the bacteria into account by \textbf{[wir sollten über diesen Punkt nochmals reden. 
            Das ist hier leicht missverständlich. Parameter sind auch kein intrazellularer Aspekt.]} \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (5) & Adhesion &
            Bacteria attract at longer distances and adhere to each other at shorter
            distances~\cite{Verwey1947,Trejo2013}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{
        This table lists aspects of cellular behaviour which we consider in our computational model.
        It is split into cellular aspects (C) which concern only an individual bacterium and
        interactions between cells (CC).
        These aspects are generalized in order to fit to as many types of rod-shaped bacteria as
        possible.
        These are the assumptions for the basic model that we are presenting here.
        In order for our methods to work, we are really only required to satisfy the assumption of
        our spatial representation (see section~\ref{subsec:spatial-representation}) and thus, these
        assumptions can also be loosened if required.
    }
\end{table}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation}
\label{subsec:spatial-representation}
This work makes use of a few but powerful assumptions concerning the mathematical representation of
the cells. In order to describe the spatial dynamics of individual cells or agents, we need to
define their position, shape and how they interact with each other.

In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
\textbf{[Dieser Satz ist leider unverständlich. Je öfter ich den Satz lese, umso weniger ist mir klar, was du sagen möchtest.]}\\

Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
\textbf{[Ebenso unklar warum du das hier an dieser Stelle sagst. Vielleicht versuchst du die Aussagen zu bündeln und geeigneter Stelle zu bringen. So ist man leider eher verwirrt.]}\\

Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
\textbf{[Hast du nicht bereits oben die Hilfsvariablen $x_i$ eingeführt?]}\\

We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
\textbf{[Nicht spekulieren was du hättest machen können, sondern einfach sagen, was du gemacht hast.]}

However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
\textbf{[Wiederholung?]}

By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and deform
plastically over longer periods of time. Our goal is to provide a model that can encapsulate these effects.
\textbf{[Das beschriebene Verhalten ist visko-elastisch. Soweit ich sehe, hast du das nicht implementiert. Könnte es sein, dass das elastisches Modell mit Wachstum äquivalent zum Maxwell-Modell ist? Eigene Antwort: im allgemeinen nein.]}\\

We discretize the (possibly non-straight) rod of a bacteria and model it by using multiple vertices
which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$. \textbf{[Was ist denn die Diskretisierungslänge? Two endpoints $x_i$ macht keinen Sinn.]}\\

Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.
\textbf{[Klingt kryptisch. "Can be used ..". Für was, wie?]}\\

With these fundamental thoughts layed out, we retain the flexibility to represent many models such
as varying shapes since we do not have to make any assumptions on the type of interactions between
two bacteria.
Even concerning the shape of bacteria, this assumption still allows us to model elliptical cells by
modifying the interaction type as well as other effects such as polarity of
interactions~\cite{Duvernoy2018}.
Furthermore, we also did not have to assume any particular mechanical dynamics of the agents, only
their spatial representation, thus leaving in particular room for plastical or elastic deformations
of the rod and other more intricate mechanisms.
\textbf{[Ich glaube, ich verstehe warum du das hier sagst. Aber das kommt nicht richtig rüber. Der Absatz ist vielleicht nicht nötig. Der Kommentar über elliptische Zellen, sollte eher unter Potential stehen?]}\\

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ have been exaggerated such that they can be identified more easily.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with no curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Mechanics}
\label{subsection:mechanical-model-mechanics}
We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness of the bacterial rod~\cite{Amir2014_2}.
As explained above, MReB will counteract the bending of the rod, resulting in a stiffening force.
In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988},
but for sake of simplicity we chose to model the bacterium as a bending beam.
\textbf{[", thus utilizing the simplest solution which can still produce the desired effects." Es ist unklar was die "desired effects" sind. ]}\\

Such a beam can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
springs of lengths $l_i$.
\textbf{[Das klingt nach einer Wiederholung von Spatial Representation. Du erklärst ziemlich umständlich. Konzentriere dich auf das wesentliche und überlege, was du eigentlich sagen möchtest.]}\\

We use the notation $\vec{c}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

@@

\begin{align}
    \partial_t^2 \vec{x}_i &= \partial_t\vec{x}_i + \sqrt{2D}\vec{\xi}_i\\
    \partial_t\vec{x}_i &= \vec{F}_{i,\text{total}} - \lambda \partial_t\vec{x}_i
\end{align}
\textbf{[Ich verstehe die Gleichungen nicht. Eine überdämpfte Langevin Gleichung bezeichnet man doch als Brownsche Bewegung, oder? Und sollte die Brownsche Gleichung nicht so lauten?
\begin{align}
   \partial_t \vec{x}_i &= \frac{D}{k_B T} \vec{F}_{i,\text{total}} + \sqrt{2D}\vec{\xi}_i
\end{align}
Die Einheiten stimmen auch überhaupt nicht. $[x_i]=$m, $[\lambda]=$kg/s, $[F]=$(kg m)/s$^2$
]}\\

where $D$ is the diffusion constant, $\vec{\xi}$ is a Wiener process and $\lambda$ a damping constant.
Together they determine the stochastic motion of the individual vertices which can arise from
recurring collisions with small particles of the surrounding medium or other minor fluctuations.

@@

Calculating the force between cells resulting  from long ranged interactions requires performing a
double integral over the length of the cells. 
To reduce the computational costs we use a approximative, simplified model Given a vertex
$\vec{x}_i$ on one cell, we determine the closest point $\vec{p}$ on the polygonal line given by the
vertices $\{\vec{w}_j\}$ of the interacting cell. 
\textbf{[Ich bin etwas verwirrt. Warum änderst du die Bezeichnung der Vertices?]}

Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

@@

The interaction between cells can be described by the Mie potential~\cite{Mie1903}.
It is the generalization of the well-known Lennard-Jones~\cite{Jones1924} potential and was designed
to describe the interactions between molecules.
The Mie potential is given by:

\begin{align}
    U(r) &= C\epsilon\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}\\
    V(r) &= \min(U(r), \beta)\theta(r-\zeta)
\end{align}

The strength of the potential is given by $\epsilon$ and the radius \textbf{[Welcher Radius?]} is
encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the radii of the two interacting
particles.
In our simulation, we use $V(r)$ to guarantee numerical regularity.
\textbf{[Beschreibe, welche Rolle $\beta$ und $\zeta$ spielen.]}

@@

\textbf{[Du hast hier von shape geredet. Das Potential ist doch immer radial-symmetrisch, oder? Vielleicht könnte es helfen, Bilder zu zeigen? Darin könnte man die Bedeutung der Parameter verdeutlichen. Wenn du schreibst, das Potential ist sehr wichtig, dann musst du viel besser begründen, warum du das Potential gewählt hast. Ich denke, du solltest das etwas abschwächen.]}

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$.
In most of our simulations, we choose to use identical lengths for all segments (i.e. $l_i=l$).
However, in order to generalize our approach we assume that individual lengths of segments could
vary. Our implemented model allows us to reduce the growth rate with increasing number of neighbours
$n\leq N$. This option provides a phenomenological approach for modelling slower growth in larger colonies but
is be toggled off by default thus removing the additional term in brackets of equation~\ref{eq:growth-rod-lengths}.

\textbf{[Ich empfinde deinen Text als uneinheitlich. Du solltest hier nicht deine Softwareimplementation beschreiben, sondern das was du tatsächlich in den Simulationen benutzt hast. Also Möglichkeiten, die du in deinem Code angelegt hast, aber für das Paper nicht verwendet hast, brauchst du nicht zu erwähnen. Das führt zu Verwirrung.]}

@@

\textbf{
    Die Verständlichkeit und die Formulierungen hier bedürfen noch der Verbesserung.
    Damit ich aber schneller durchkomme, verschieben wir das auf später.
}

@@

\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar.]}

@@

%###################################################################################################
\section{Application 2: \textit{E.Coli.} growing on Agar}
\label{section:application-2}
To show the usefulness of our previously developed methods
(see section~\ref{section:parameter-estimation}), we will use them to estimate the parameters
individual growing \textit{E.Coli} using a sequence of microscopic images.
The data we use~\cite{https://doi.org/10.3203/iwf/k-129} (see
section~\ref{section:data-code-availability}) has been automatically segmented with the
omnipose~\cite{Cutler2022} algorithm by mean of which the necessary cell masks are obtained.
